Alternative Methods for Implementing Explicit and Finding Implicit Sharing in embedded DSLs

### 2022-12-14 paper/paper.tex
background done

Implicit sharing via hash-consing prevents duplication in the resulting DAG, but
unfortunately doesn't prevent redundant computation. Consider the following
equivalent attempt at using Haskell's built-in sharing in the finally tagless DSL
\begin{minted}{haskell}
dslSharing :: Exp Graph -> Exp Graph -> Exp Graph
dslSharing x y =
  let
    z = add x y
  in add z z
\end{minted}
Note {\bf z} is a wrapper around a State monad. Recall the implementation of
{\bf add} via hash consing
\begin{minted}{haskell}
add e1 e2 = Graph (do
                   h1 <- unGraph e1
                   h2 <- unGraph e2
                   hashcons $ NAdd h1 h2)
\end{minted}
The values {\bf h1} and {\bf h2} need to be explicitely evaluated through the
State monad, meaning even if {\bf e1} and {\bf e2} are the same shared Haskell
value, their underlying computations will be performed twice. Hash-consing will
prevent these redundencies from appearing in the resulting DAG, the the entire
unshared AST wil still be traversed, performing a hash-cons on each node.

TODO describe add-chains example

TODO translate to kiselyov's language
\begin{minted}{haskell}
addChains :: forall repr. CoreISA repr => (repr VR, repr VR, repr VR, repr VR) -> (repr VR, repr VR, repr VR, repr VR)
addChains =
      add2 . add1
    . add2 . add1
    . add2 . add1
    . add2 . add1
    . add2 . add1
\end{minted}

\subsection{Explicit Sharing and Limitations}
TODO

>>>>

Implicit sharing via hash-consing prevents duplication in the resulting DAG, but
unfortunately doesn't prevent redundant computation. Consider the following
equivalent attempt at using Haskell's built-in sharing in the finally tagless DSL
\begin{minted}{haskell}
dslSharing :: Exp Graph -> Exp Graph -> Exp Graph
dslSharing x y =
  let
    z = add x y
  in add z z
\end{minted}
Note {\bf z} is a wrapper around a State monad. Recall the implementation of
{\bf add} via hash consing
\begin{minted}{haskell}
add e1 e2 = Graph (do
                   h1 <- unGraph e1
                   h2 <- unGraph e2
                   hashcons $ NAdd h1 h2)
\end{minted}
The values {\bf h1} and {\bf h2} need to be explicitely evaluated through the
State monad, meaning even if {\bf e1} and {\bf e2} are the same shared Haskell
value, their underlying computations will be performed twice. Hash-consing will
prevent these redundencies from appearing in the resulting DAG, but the entire
unshared AST wil still be traversed, performing a hash-cons on each node.

Consider a chain of {\bf add}'s  with sharing, for example
\begin{minted}{haskell}
addChains :: Exp repr => Expr Int -> Expr Int
addChains x0 = let
    x1 = add x0 x0
    x2 = add x1 x1
    ...
  in xn
\end{minted}

\begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth]{figs/hashcons.png}
  \caption{Number of hashcons to add operations performed} \label{fig:hashcons}
\end{figure}
As you can see from figure \ref{fig:hashcons}, this code will perform approximately
$2^{n+1}$ hashcons operations, where n is the number of {\bf add}'s.

\subsection{Explicit Sharing and Limitations}

\cite{kiselyov:sharing} acknowledges the amount of computation with hash-consing
can get out of control, and proposes an ad-hoc solution, explicit sharing via a
custom let construct
\begin{minted}{haskell}
class ExpLet repr where
  let_ :: repr a -> (repr a -> repr b) -> repr b
instance ExpLet Graph where
  let_ e f = Graph (do x <- unGraph e
                     unGraph $ f (Graph (return x)))

addChains x =
  let_ x (\x0 ->
  let_ (add x0 x0)  (\x1 ->
  let_ (add x1 x1)  (\x2 ->
   ...
  )))
\end{minted}
This makes the code a bit clunky and adds an extra burden on the DSL
writer, but it prevents unnecessary hash-consing in our example. However the
method has it's limitations, suppose we want to extend our DSL to support
functions that return multiple outputs, such as tuples or container types like
lists (for example \mintinline{haskell}{vadd :: (repr Int,repr Int) -> (repr Int,repr
  Int)}). First of all, we need to implement different versions of the custom let
construct to correspond to the number of outputs
\begin{minted}{haskell}
class ExpLet repr where
  let_ :: repr a -> (repr a -> repr b) -> repr b
  let_ 2 :: repr a -> (repr a -> (repr b,repr c)) -> (repr b,repr c)

instance ExpLet repr where
   ...
   let_2 e f = let
       g0 = Graph (do x <- unGraph e
                        let (o0,o1) = f (Graph (return x))
                        unGraph o0)
       g1 = Graph (do x <- unGraph e
                        let (o0,o1) = f (Graph (return x))
                        unGraph o1)
       in (g0,g1)
\end{minted}
If need be, its possible to enumerate custom let instances for every amount of
outputs or container types we would need, we could even use template Haskell to
accomplish this. However, this custom let construct now has a new source of
redundancy in its outputs. Each output it returns will now have to individually
evaluate it's input, so a chain of DSL functions that output 2 or more values
will suffer from the same exponential scaling of hashcons.

One solution to this issue is to integrate container types such as tuples and lists
into the DSL language. However doing this will essentially make the DSL ``less
embedded'', manipulating tuple values will be cumbersome constantly requiring
calls to custom implementations of {\bf fst}/{\bf snd} etc. And for lists you'll
lose all access to built-in Haskell list functionality.
